Computing in the Field of Rationals

### 2012-06-26 rational_ext.tex
_

%subcode-inline{bnf-inline} name langRev
%! swap+ = \mathit{swap}^+
%! swap* = \mathit{swap}^*
%! dagger =  ^{\dagger}
%! assocl+ = \mathit{assocl}^+
%! assocr+ = \mathit{assocr}^+
%! assocl* = \mathit{assocl}^*
%! assocr* = \mathit{assocr}^*
%! identr* = \mathit{uniti}
%! identl* = \mathit{unite}
%! dist = \mathit{distrib}
%! factor = \mathit{factor}
%! eta = \eta
%! eps = \epsilon
%! eta+ = \eta^+
%! eps+ = \epsilon^+
%! eta* = \eta^{\times}
%! eps* = \epsilon^{\times}
%! trace+ = trace^+
%! trace* = trace^{\times}
%! ^^^ = ^{-1}
%! (o) = \circ
%! (;) = \fatsemi
%! (*) = \times
%! (+) = +
%! LeftP = L^+
%! RightP = R^+
%! LeftT = L^{\times}
%! RightT = R^{\times}
%! alpha = \alpha
%! bool = \textit{bool}
%! color = \textit{color}
%! Gr = G

%subcode-inline{bnf-inline} regex \{\{(((\}[^\}])|[^\}])*)\}\} name main include langRev
%! Gx = \Gamma^{\times}
%! G = \Gamma
%! [] = \Box
%! |-->* = \mapsto^{\asterix}
%! |-->> = \mapsto_{\ggg}
%! |--> = \mapsto
%! <--| = \mapsfrom
%! |- = \vdash
%! <><> = \approx
%! ==> = \Longrightarrow
%! <== = \Longleftarrow
%! <=> = \Longleftrightarrow
%! <-> = \leftrightarrow
%! ~> = \leadsto
%! -o+ = \multimap^{+}
%! -o* = \multimap^{\times}
%! -o = \multimap
%! ::= = &::=&
%! /= = \neq
%! @@ = \mu
%! [^ = \lceil
%! ^] = \rceil
%! forall = \forall
%! exists = \exists
%! empty = \epsilon
%! Pi = \Pi
%! Pi0 = \Pi^{o}
%! PiEE* = \Pi^{\eta\epsilon}_{*}
%! PiEE+ = \Pi^{\eta\epsilon}_{+}
%! PiEE = \Pi^{\eta\epsilon}
%! CatSet = \textbf{Set}
%! theseus = Theseus
%! sqrt(x) = \sqrt{#x}
%! surd(p,x) = \sqrt[#p]{#x}
%! inv(x) = \frac{1}{#x}
%! frac(x,y) = \frac{#x}{#y}
%! * = \times

@@

The usual interpretation of {{b1*b2}} we have both a value of type
{{b1}} and a value of type {{b2}}. This interpretation is maintained
in the presence of fractionals. Hence an {{eta* : 1 <-> (1/b) * b}} is
to be viewed as a fission point for a value of type {{b}} and its
multiplicative inverse {{1/b}}. Operationally, this corresponds to the
creation of two values {{alpha}} and {{1/alpha}} where {{alpha}} is a
fresh logic variable. The operator {{eps*}} then becomes a unification
site for these logic variables:

@@

@@

% Note that this language is not logically reversible.

>>>>

%subcode-inline{bnf-inline} name langRev
%! swap+ = \mathit{swap}^+
%! swap* = \mathit{swap}^*
%! dagger =  ^{\dagger}
%! assocl+ = \mathit{assocl}^+
%! assocr+ = \mathit{assocr}^+
%! assocl* = \mathit{assocl}^*
%! assocr* = \mathit{assocr}^*
%! identr* = \mathit{uniti}
%! identl* = \mathit{unite}
%! dist = \mathit{distrib}
%! factor = \mathit{factor}
%! eta = \eta
%! eps = \epsilon
%! eta+ = \eta^+
%! eps+ = \epsilon^+
%! eta* = \eta^{\times}
%! eps* = \epsilon^{\times}
%! trace+ = trace^+
%! trace* = trace^{\times}
%! (o) = \circ
%! (;) = \fatsemi
%! (*) = \times
%! (+) = +
%! LeftP = L^+
%! RightP = R^+
%! LeftT = L^{\times}
%! RightT = R^{\times}
%! alpha = \alpha
%! bool = \textit{bool}
%! color = \textit{color}
%! Gr = G

%subcode-inline{bnf-inline} regex \{\{(((\}[^\}])|[^\}])*)\}\} name main include langRev
%! Gx = \Gamma^{\times}
%! G = \Gamma
%! [] = \Box
%! |-->* = \mapsto^{\asterix}
%! |-->> = \mapsto_{\ggg}
%! |--> = \mapsto
%! <--| = \mapsfrom
%! |- = \vdash
%! <><> = \approx
%! ==> = \Longrightarrow
%! <== = \Longleftarrow
%! <=> = \Longleftrightarrow
%! <-> = \leftrightarrow
%! ~> = \leadsto
%! -o+ = \multimap^{+}
%! -o* = \multimap^{\times}
%! -o = \multimap
%! ::= = &::=&
%! /= = \neq
%! @@ = \mu
%! [^ = \lceil
%! ^] = \rceil
%! {| = \{
%! |} = \}
%! ^^^ = ^{\dagger}
%! elem = \in
%! forall = \forall
%! exists = \exists
%! empty = \epsilon
%! Pi = \Pi
%! Pi0 = \Pi^{o}
%! PiEE* = \Pi^{\eta\epsilon}_{*}
%! PiEE+ = \Pi^{\eta\epsilon}_{+}
%! PiEE = \Pi^{\eta\epsilon}
%! CatSet = \textbf{Set}
%! theseus = Theseus
%! sqrt(x) = \sqrt{#x}
%! surd(p,x) = \sqrt[#p]{#x}
%! inv(x) = \frac{1}{#x}
%! frac(x,y) = \frac{#x}{#y}
%! * = \times

@@

The usual interpretation of {{b1*b2}} we have both a value of type
{{b1}} and a value of type {{b2}}. This interpretation is maintained
in the presence of fractionals. Hence an {{eta* : 1 <-> (1/b) * b}} is
to be viewed as a fission point for a value of type {{b}} and its
multiplicative inverse {{1/b}}. 
%% Operationally, this corresponds to the
%% creation of two values {{alpha}} and {{1/alpha}} where {{alpha}} is a
%% fresh logic variable. The operator {{eps*}} then becomes a unification
%% site for these logic variables:

@@

\roshan{Maybe replace the annihilation example with union and
  difference.}

@@

In the discourse below we use {{c v |-->* v'}} as a shorthand for the
execution sequence {{<c,v,[]>|-->*{[c,v',[] ]} }}.

\begin{definition}[eval(c,v)]

Given a combinator {{c:b1<->b2}} and a value {{v:b1}} we define
{{eval(c, v)}} to be the set of values {{v':b}} such that there is
small-step reduction sequence {{c v |-->* v'}}.

$eval(c, v) = \{v'~|~ c~ v\mapsto^{\asterix} v' \} $
\end{definition}

The above definition implies that combinators {{c:b1<->b2}} can be
indetified by relations between the sets {{b1}} and {{b2}}, i.e. 
{{c:b1<->b2 = {| (v, v') | v:b2, v':b2 and c v |-->* v'|} }}.
The definition of equality between combinators is extended reflect the
fact that combinators denote relations.

\begin{definition}[c1 = c2]
We say that {{c1 = c2}}, for combinators {{c1:b1<->b2}} and
{{c2:b1<->b2}} iff they denote the same relation between sets {{b1}}
and {{b2}}. Equivalently,

{{forall v:b1. eval(c1, v) = eval(c2, v)}}
\end{definition}

\begin{proposition}[Logical Reversibility]
\label{prop:logrev2} ~

{{c v |-->* v'}} iff {{ c{dagger} v' |-->* v}}
\end{proposition}

Logical reversibility of {{PiEE*}} implies that {{c^{dagger} }}
denotes the inverse relation that {{c}} denotes, i.e.
{{c^^^={|(v',v)~|~(v,v') elem c|} }}.  Operationally, for any
execution path from {{v}} to {{v'}} in {{c}}, {{c{dagger} }} has an
execution path from {{v'}} to {{v}}.

\roshan{This is an important question -- is {{PiEE*}} still
  bimonoidal? i.e. do all the coherence conditions on isomorphisms
  naturally extend to relations. I suspect yes.}
